---
title: "Mouse PR degeneration literature"
subtitle: "Statistical modeling using multinomial regression"
author: "Luca Della Santina, UCSF (Luca.DellaSantina@ucsf.edu)"
date: "Updated: `r Sys.time()`"
output:
  html_document:
    highlight: haddock
    theme: default
    code_folding: hide
    df_print: paged
    toc: yes
    toc_depth: '4'
    toc_float:
      collapsed: no
      smooth_scroll: yes
---
  
# Summary
  
  Multinomial regression analysis of PR degeneration literature compiled by JL.
  Purpose of the model is to understand which factors are important in determining a specific outcome (Stabilty vs Degeneration vs Compensation)
  

  
# Preamble
```{r preamble}
require(car)
require(reshape2)
require(foreign)
require(nnet)
require(ggplot2)
library(tidyverse)
require(ztable)
```

# Load and format data
Load the formatted data table provided by JL

```{r load the data2, warning=FALSE, message=FALSE}
library(readr)
df <- read_csv("./PhotoreceptorLiterature_20201015.csv", 
    col_types = cols(Outcome = col_factor(levels = c("Stability", 
        "Degeneration", "Compensation")), 
        Method = col_factor(levels = c("Str","Fun")), 
        Perturbation = col_factor(levels = c("Dev","Mat")), 
        RodCone = col_factor(levels = c("Rod","Cone", "RodCone")), 
        Rod_YN = col_factor(levels = c("NotRod","Rod")), 
        Cone_YN = col_factor(levels = c("NotCone","Cone")), 
        Phototransduction_YN = col_factor(levels = c("NotPT","Phototransduction")), 
        Signaling_YN = col_factor(levels = c("NotSig","Signaling")), 
        Trafficking_YN = col_factor(levels = c("NotTff","Trafficking")), 
        Metabolism_YN = col_factor(levels = c("Notmet","Meta")), 
        ImmuneDNA_YN = col_factor(levels = c("NotImmune","Immune")), 
        TF_YN = col_factor(levels = c("NotTF","TF")), 
        ManipulationFull = col_factor(levels = c("Phototransduction","Signaling", "Trafficking", "Meta","Immune", "TF")),
        Manipulation = col_factor(levels = c("Phototransduction", "Signaling", "Trafficking", "Others"))))
#View(df)
```
# Frequency tables

Contingency tables for each factor, expressed first as absolute count then as percent frequency for each factor's level

```{r Exploratory tables}

writeLines("Method of investigation (Str= Structure, Fun=Function)\n\nFrequency count:")
ftable(df$Method, df$Outcome)
writeLines("\nRelative Frequency %:")
prop.table(ftable(df$Method, df$Outcome))

writeLines("\nType of perturbation (Dev=Developmental, Mature)\n\nFrequency count:")
ftable(df$Perturbation, df$Outcome)
writeLines("\nRelative Frequency %:")
prop.table(ftable(df$Perturbation, df$Outcome))

writeLines("\nType of photoreceptor affected (Rod, Cone, Rod+Cone)\n\nFrequency count:")
ftable(df$RodCone, df$Outcome)
writeLines("\nRelative Frequency %:")
prop.table(ftable(df$RodCone, df$Outcome))

writeLines("\nType of Manipulation\n\nFrequency count:")
ftable(df$Manipulation, df$Outcome)
writeLines("\nRelative Frequency %:")
prop.table(ftable(df$Manipulation, df$Outcome))

```

# Exploratory Plotting

Some factors don't have enough variability (<5% or >95%), therefore they will not be excluded in the model.

```{r Exploratory plotting}

#par(mfrow = c(3, 5), mar=c(2,2,2,2))

# Categorical
plot(df$Outcome, df$Method, xlab = "Outcome", ylab = "Method of investigation")
plot(df$Outcome, df$Perturbation, xlab = "Outcome", ylab = "Developmental vs Mature Perturbation")
plot(df$Outcome, df$RodCone, xlab = "Outcome", ylab = "Type of photoreceptor involved")

plot(df$Outcome, df$Phototransduction_YN, xlab = "Outcome", ylab = "Phototransduction")
plot(df$Outcome, df$Signaling_YN, xlab = "Outcome", ylab = "Signaling")
plot(df$Outcome, df$Trafficking_YN, xlab = "Outcome", ylab = "Trafficking")
plot(df$Outcome, df$Metabolism_YN, xlab = "Outcome", ylab = "Metabolism")
plot(df$Outcome, df$ImmuneDNA_YN, xlab = "Outcome", ylab = "Immune DNA")
plot(df$Outcome, df$TF_YN, xlab = "Outcome", ylab = "Transcription Factors")

plot(df$Outcome, df$Manipulation, xlab = "Outcome", ylab = "Manipulation (PT/Signaling/Trafficking/Others")

# Numerical
plot(df$Outcome, df$NormalizedAge, xlab = "Outcome", ylab = "Age (Normalized by development time)")
plot(df$Outcome, df$NormalizedInterval, xlab = "Outcome", ylab = "Interval (Normalized by development time)")
plot(df$Outcome, df$EntirePR, xlab = "Outcome", ylab = "Entire PR")
plot(df$Outcome, df$SpecificPR, xlab = "Outcome", ylab = "Specific PR")
```

# Model 1: All parameters
## Multinomial Regression Model

Multinomial regression (nnet:multinom), it is a log-linear model for which the resulting coefficients reflect the ratio of probability

```{r model 1}

df$Outcome<-relevel(df$Outcome, ref="Stability")

mdl <- multinom(Outcome ~ Method + RodCone +  Manipulation + NormalizedAge + NormalizedInterval + EntirePR + SpecificPR, data = df)
summary(mdl)
```

## ANOVA

Parameters significant for the model are: Method of investigation, Rod/Cone affected, and Entire PR affected.

```{R ANOVA model 1}
Anova(mdl)
```

# Model 2: Method + RodCone
## Multinomial Regression Model

Multinomial regression (nnet:multinom), it is a log-linear model for which the resulting coefficients reflect the ratio of probability
Removed the Manipulation parameter because not significant in the model

```{r model 2}

df$Outcome<-relevel(df$Outcome, ref="Stability")

mdl <- multinom(Outcome ~ Method + RodCone + NormalizedAge + NormalizedInterval + EntirePR + SpecificPR, data = df)
summary(mdl)
```

## ANOVA

Parameters significant for the model are: Method of investigation, Rod/Cone affected, and Entire PR affected.

```{R anova model 2}
Anova(mdl)
```
## Wald-test (2-tailed z test)
```{r Z-test}
z <- summary(mdl)$coefficients/summary(mdl)$standard.errors
p <- (1 - pnorm(abs(z), 0, 1)) * 2
p
```

## Relative risk

```{r extract coefficients}
writeLines("Ratio of the probability of choosing one outcome category over the reference category.\nOften referred as relative risk (and sometimes referred to as odds).\nCalculated by extracting the coefficients from the model and exponentiating\n")
exp(coef(mdl))

#writeLines("\nPredicted probabilities for the source data (head of table)")
#head(fitted(mdl))
```

## Predicted probabilities
### Categorical Parameters

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ztable)
library(magrittr)
options(ztable.type = "html")
```

```{r predicted probabilities, results='asis'}
dsimStr <- data.frame(Method = "Str", 
                   RodCone = levels(df$RodCone), 
                   NormalizedAge = mean(df$NormalizedAge), 
                   NormalizedInterval = mean(df$NormalizedInterval), 
                   EntirePR = mean(df$EntirePR), 
                   SpecificPR = mean(df$SpecificPR))
predStr <- predict(mdl, newdata = dsimStr, "probs")

dsimFun <- data.frame(Method = "Fun", 
                   RodCone = levels(df$RodCone), 
                   NormalizedAge = mean(df$NormalizedAge), 
                   NormalizedInterval = mean(df$NormalizedInterval), 
                   EntirePR = mean(df$EntirePR), 
                   SpecificPR = mean(df$SpecificPR))
predFun <- predict(mdl, newdata = dsimFun, "probs")

predStrFun <- rbind(predStr, predFun)
rgroup=c("Structure","Function")
n.rgroup=c(3,3)
dfPred <- as.data.frame(predStrFun)
row.names(dfPred) <- NULL
dfPred <- roundDf(dfPred, digits=2)
Photoreceptor <- c(levels(df$RodCone),levels(df$RodCone)) 
dfPred <- cbind(Photoreceptor, dfPred)
z <- ztable(dfPred) %>%
  addrgroup(rgroup=rgroup,n.rgroup=n.rgroup,cspan.rgroup=1) %>%
  print(caption = "Predicted outcome probability by method and photoreceptor type")
```

### Age

Predicts outcome based on age (normalized by developmental time) at which the pathology starts

```{R Predicted Prob Age}
dNormalizedAge <- data.frame(Method = rep(levels(df$Method), each = length(unique(df$NormalizedAge))*length(levels(df$RodCone))), 
                   RodCone = rep(levels(df$RodCone), each = length(unique(df$NormalizedAge))), 
                   NormalizedAge = rep(sort(unique(df$NormalizedAge)), length(levels(df$RodCone))), 
                   NormalizedInterval = mean(df$NormalizedInterval), 
                   EntirePR = mean(df$EntirePR), 
                   SpecificPR = mean(df$SpecificPR))

# Store the predicted probabilities for each value of RodCone and NormalizedAge and append as last 3 columns to data frame (last three columns)
pred <- predict(mdl, newdata = dNormalizedAge, type = "probs", se = TRUE)
pp <- cbind(dNormalizedAge, pred)

# calculate the mean probabilities within each level of RodCone 
#by(pp[, (ncol(pp)-2):ncol(pp)], pp$RodCone, colMeans)

# Collapse the data frame by outcome
lpp <- melt(pp, id.vars = c("RodCone", "NormalizedAge", "Method"), measure.vars = levels(df$Outcome), value.name = "Probability")

# Plot predicted probabilities across NormalizedAge values for each level of RodCone facetted by Outcome and Method of Investigation
ggplot(lpp, aes(x = NormalizedAge, y = Probability, colour = RodCone)) +
  geom_line() + facet_grid(variable ~ Method, scales = "free_x") +
  ggtitle("Predicted Outcome by Age (normalized by developmental time)") + 
  theme(plot.title = element_text(hjust = 0.5)) + ylim(0,1)

# Plot predicted probabilities across NormalizedAge values for each level of Outcome facetted by RodCone and Method of Investigation
ggplot(lpp, aes(x = NormalizedAge, y = Probability, colour = variable)) +
  geom_line() + facet_grid(RodCone ~ Method, scales = "free_x") +
  ggtitle("Predicted Outcome by Age (normalized by developmental time)") + 
  theme(plot.title = element_text(hjust = 0.5)) + ylim(0,1)
```

### Interval

Predicts outcome based on time interval (normalized by developmental time) during which the pathology develops

```{R Predicted Prob NormalizedInterval}
dNormalizedInterval <- data.frame(Method = rep(levels(df$Method), each = length(unique(df$NormalizedInterval))*length(levels(df$RodCone))), 
                   RodCone = rep(levels(df$RodCone), each = length(unique(df$NormalizedInterval))), 
                   NormalizedAge =  mean(df$NormalizedAge),
                   NormalizedInterval = rep(sort(unique(df$NormalizedInterval)), length(levels(df$RodCone))), 
                   EntirePR = mean(df$EntirePR), 
                   SpecificPR = mean(df$SpecificPR))

# Store the predicted probabilities for each value of RodCone and D50 and append as last 3 columns to dataframe (columns 9, 10, 11)
pp <- cbind(dNormalizedInterval, predict(mdl, newdata = dNormalizedInterval, type = "probs", se = TRUE))

# calculate the mean probabilities within each level of RodCone 
#by(pp[, (ncol(pp)-2):ncol(pp)], pp$RodCone, colMeans)

# Collapse the data frame by outcome
lpp <- melt(pp, id.vars = c("RodCone", "NormalizedInterval", "Method"), measure.vars = levels(df$Outcome), value.name = "Probability")

# Plot predicted probabilities across NormalizedInterval values for each level of RodCone facetted by Outcome and Method of investigation
ggplot(lpp, aes(x = NormalizedInterval, y = Probability, colour = RodCone)) + 
  geom_line() + facet_grid(variable ~ Method, scales = "free_x") + 
  ggtitle("Predicted Outcome by Interval (normalized by developmental time)") + 
  theme(plot.title = element_text(hjust = 0.5)) + ylim(0,1)

# Plot predicted probabilities across NormalizedInterval values for each level of Outcome facetted by RodCone and Method of investigation
ggplot(lpp, aes(x = NormalizedInterval, y = Probability, colour = variable)) + 
  geom_line() + facet_grid(RodCone ~ Method, scales = "free_x") + 
  ggtitle("Predicted Outcome by Interval (normalized by developmental time)") + 
  theme(plot.title = element_text(hjust = 0.5)) + ylim(0,1)
```

### Entire PR

Predict probability by total amount of photoreceptors affected, assuming the following distribution:
Cones: between 0 and 3%, Rods: between 0 and 97%

```{R Predicted Prob Entire PR}
percCone = seq(0, 3, by=0.5)
percRod = seq(0, 97, by=0.5)
percRodCone = seq(0, 100, by=0.5)

dConePR = data.frame(Method = rep(levels(df$Method), each = length(percCone)), 
                   RodCone = rep("Cone", each = length(percCone)), 
                   NormalizedAge =  mean(df$NormalizedAge),
                   NormalizedInterval = mean(df$NormalizedInterval), 
                   EntirePR = percCone, 
                   SpecificPR = mean(df$SpecificPR))

dRodPR = data.frame(Method = rep(levels(df$Method), each = length(percRod)), 
                   RodCone = rep("Rod", each = length(percRod)), 
                   NormalizedAge =  mean(df$NormalizedAge),
                   NormalizedInterval = mean(df$NormalizedInterval), 
                   EntirePR = percRod, 
                   SpecificPR = mean(df$SpecificPR))

dRodConePR = data.frame(Method = rep(levels(df$Method), each = length(percRodCone)), 
                   RodCone = rep("RodCone", each = length(percRodCone)), 
                   NormalizedAge =  mean(df$NormalizedAge),
                   NormalizedInterval = mean(df$NormalizedInterval), 
                   EntirePR = percRodCone, 
                   SpecificPR = mean(df$SpecificPR))

dEntirePR = rbind(dConePR, dRodPR, dRodConePR)

#dEntirePR <- data.frame(Method = rep(levels(df$Method), each = length(unique(df$EntirePR))*length(levels(df$RodCone))), 
#                   RodCone = rep(levels(df$RodCone), each = length(unique(df$EntirePR))), 
#                   NormalizedAge =  mean(df$NormalizedAge),
#                   NormalizedInterval = mean(df$NormalizedInterval), 
#                   EntirePR = rep(sort(unique(df$EntirePR)), length(levels(df$RodCone))), 
#                   SpecificPR = mean(df$SpecificPR))

# Store the predicted probabilities for each value of RodCone and D50 and append as last 3 columns to dataframe (columns 9, 10, 11)
pp <- cbind(dEntirePR, predict(mdl, newdata = dEntirePR, type = "probs", se = TRUE))

# calculate the mean probabilities within each level of RodCone 
#by(pp[, (ncol(pp)-2):ncol(pp)], pp$RodCone, colMeans)

# Collapse the data frame by outcome
lpp <- melt(pp, id.vars = c("RodCone", "EntirePR", "Method"), measure.vars = levels(df$Outcome), value.name = "Probability")

# Plot predicted probabilities across EntirePR values for each level of RodCone facetted by Outcome and Method
ggplot(lpp, aes(x = EntirePR, y = Probability, colour = RodCone)) + 
  geom_line() + facet_grid(variable ~ Method, scales = "free")  + 
  ggtitle("Predicted Outcome by Percentage of Enrire Photoreceptors Affected") + 
  theme(plot.title = element_text(hjust = 0.5))  + ylim(0,1)

# Plot predicted probabilities across EntirePR values for each level of Outcome facetted by RodCone and Method
ggplot(lpp, aes(x = EntirePR, y = Probability, colour = variable)) + 
  geom_line() + facet_grid(RodCone ~ Method, scales = "free")  + 
  ggtitle("Predicted Outcome by Percentage of Enrire Photoreceptors Affected") + 
  theme(plot.title = element_text(hjust = 0.5))  + ylim(0,1)
```

### Specific PR

Predict probability of each outcome by the amount of scpecific photoreceptors affected.

```{R Predicted Prob Specific PR}
percRodCone = seq(0, 100, by=0.5)
typePR = c("Rod", "Cone")

dSpecificPR <- data.frame(Method = rep(levels(df$Method), each = length(percRodCone)*length(typePR)), 
                   RodCone = rep(typePR, each = length(percRodCone)), 
                   NormalizedAge =  mean(df$NormalizedAge),
                   NormalizedInterval = mean(df$NormalizedInterval), 
                   EntirePR =  mean(df$EntirePR),
                   SpecificPR = rep(percRodCone, length(levels(df$Method))*length(typePR)))

# Store the predicted probabilities for each value of RodCone and D50 and append as last 3 columns to dataframe (columns 9, 10, 11)
pp <- cbind(dSpecificPR, predict(mdl, newdata = dSpecificPR, type = "probs", se = TRUE))

# calculate the mean probabilities within each level of RodCone 
#by(pp[, (ncol(pp)-2):ncol(pp)], pp$RodCone, colMeans)

# Collapse the data frame by outcome
lpp <- melt(pp, id.vars = c("RodCone", "SpecificPR", "Method"), measure.vars = levels(df$Outcome), value.name = "Probability")

# Plot predicted probabilities across SpecificPR values for each level of RodCone facetted by Outcome and Method
ggplot(lpp, aes(x = SpecificPR, y = Probability, colour = RodCone)) + 
  geom_line() + facet_grid(variable ~ Method, scales = "free")  + 
  ggtitle("Predicted Outcome by Percentage of Specific Photoreceptors Affected") + 
  theme(plot.title = element_text(hjust = 0.5))  + ylim(0,1)

# Plot predicted probabilities across SpecificPR values for each level of Outcome facetted by RodCone and Method
ggplot(lpp, aes(x = SpecificPR, y = Probability, colour = variable)) + 
  geom_line() + facet_grid(RodCone ~ Method, scales = "free")  + 
  ggtitle("Predicted Outcome by Percentage of Specific Photoreceptors Affected") + 
  theme(plot.title = element_text(hjust = 0.5))  + ylim(0,1)
```

# Model 3: RodCone
## Multinomial Regression Model

Multinomial regression (nnet:multinom), it is a log-linear model for which the resulting coefficients reflect the ratio of probability
Removed the Manipulation parameter because not significant in the model and the Method of investigtion to simplify prediction

```{r model 3}

df$Outcome<-relevel(df$Outcome, ref="Stability")

mdl <- multinom(Outcome ~ RodCone + NormalizedAge + NormalizedInterval + EntirePR + SpecificPR, data = df)
summary(mdl)
```

## ANOVA

Parameters significant for the model are: Rod/Cone affected and Entire PR affected.

```{R anova model 3}
Anova(mdl)
```
## Wald-test (2-tailed z test)
```{r Z-test model 3}
z <- summary(mdl)$coefficients/summary(mdl)$standard.errors
p <- (1 - pnorm(abs(z), 0, 1)) * 2
p
```

## Relative risk

```{r extract coefficients model 3}
writeLines("Ratio of the probability of choosing one outcome category over the reference category.\nOften referred as relative risk (and sometimes referred to as odds).\nCalculated by extracting the coefficients from the model and exponentiating\n")
exp(coef(mdl))

#writeLines("\nPredicted probabilities for the source data (head of table)")
#head(fitted(mdl))
```

## Predicted probabilities
### Categorical Parameters

```{r predicted probabilities model 3, results='asis'}
dsim <- data.frame(RodCone = levels(df$RodCone), 
                   NormalizedAge = mean(df$NormalizedAge), 
                   NormalizedInterval = mean(df$NormalizedInterval), 
                   EntirePR = mean(df$EntirePR), 
                   SpecificPR = mean(df$SpecificPR))
pred <- predict(mdl, newdata = dsim, "probs")

dfPred <- as.data.frame(pred)
row.names(dfPred) <- NULL
dfPred <- roundDf(dfPred, digits=2)
Photoreceptor <- c(levels(df$RodCone)) 
dfPred <- cbind(Photoreceptor, dfPred)
z <- ztable(dfPred) %>%
  print(caption = "Predicted outcome probability by photoreceptor type")
```

### Age

Predicts outcome based on age (normalized by developmental time) at which the pathology starts

```{R Predicted Prob Age model 3}
# Distribution of data (cone = 0 to 1.428, rod = 0 to 1.428, rod/cone 0 to 2.666)
percCone = seq(0, 3.0, by=0.5)
percRod = seq(0, 3.0, by=0.5)
percRodCone = seq(0, 3.0, by=0.5)

dConePR <-data.frame(RodCone = rep("Cone", each = length(percCone)), 
                   NormalizedAge =  percCone,
                   NormalizedInterval = mean(df$NormalizedInterval), 
                   EntirePR =  mean(df$EntirePR), 
                   SpecificPR = mean(df$SpecificPR))

dRodPR <-data.frame(RodCone = rep("Rod", each = length(percRod)), 
                   NormalizedAge =  percRod,
                   NormalizedInterval = mean(df$NormalizedInterval), 
                   EntirePR = mean(df$EntirePR), 
                   SpecificPR = mean(df$SpecificPR))
  
dRodConePR <- data.frame(RodCone = rep("RodCone", each = length(percRodCone)), 
                   NormalizedAge =  percRodCone,
                   NormalizedInterval = mean(df$NormalizedInterval), 
                   EntirePR = mean(df$EntirePR), 
                   SpecificPR = mean(df$SpecificPR))

dNormalizedAge <- rbind(dConePR, dRodPR, dRodConePR)
# Store the predicted probabilities for each value of RodCone and NormalizedAge and append as last 3 columns to data frame (last three columns)
pred <- predict(mdl, newdata = dNormalizedAge, type = "probs", se = TRUE)
pp <- cbind(dNormalizedAge, pred)

# calculate the mean probabilities within each level of RodCone 
#by(pp[, (ncol(pp)-2):ncol(pp)], pp$RodCone, colMeans)

# Collapse the data frame by outcome
lpp <- melt(pp, id.vars = c("RodCone", "NormalizedAge"), measure.vars = levels(df$Outcome), value.name = "Probability")

# Plot predicted probabilities across NormalizedAge values for each level of RodCone facetted by Outcome
ggplot(lpp, aes(x = NormalizedAge, y = Probability, colour = RodCone)) +
  geom_line() + facet_grid(variable ~ ., scales = "free_x") +
  ggtitle("Predicted Outcome by Age (normalized by developmental time)") + 
  theme(plot.title = element_text(hjust = 0.5)) + ylim(0,1)

# Collapse the data frame by RodCone
lpp <- melt(pp, id.vars = c("RodCone", "NormalizedAge"), measure.vars = levels(df$Outcome), value.name = "Probability")

# Plot predicted probabilities across NormalizedAge values for each level of Outcome facetted by RodCone
ggplot(lpp, aes(x = NormalizedAge, y = Probability, colour = variable)) +
  geom_line() + facet_grid(RodCone ~ ., scales = "free_x") +
  ggtitle("Predicted Outcome by Age (normalized by developmental time)") + 
  theme(plot.title = element_text(hjust = 0.5)) + ylim(0,1)

```

### Interval

Predicts outcome based on time interval (normalized by developmental time) during which the pathology develops

```{R Predicted Prob NormalizedInterval model 3}
# Distribution of data (cone = 0 to 24.28, rod = 0 to 40, rod/cone 0 to 34.28)
percCone = seq(0, 40, by=2)
percRod = seq(0, 40, by=2)
percRodCone = seq(0, 40, by=2)

dConePR <-data.frame(RodCone = rep("Cone", each = length(percCone)), 
                   NormalizedAge =  mean(df$NormalizedAge),
                   NormalizedInterval = percCone, 
                   EntirePR =  mean(df$EntirePR), 
                   SpecificPR = mean(df$SpecificPR))

dRodPR <-data.frame(RodCone = rep("Rod", each = length(percRod)), 
                   NormalizedAge =  mean(df$NormalizedAge),
                   NormalizedInterval = percRod, 
                   EntirePR = mean(df$EntirePR), 
                   SpecificPR = mean(df$SpecificPR))
  
dRodConePR <- data.frame(RodCone = rep("RodCone", each = length(percRodCone)), 
                   NormalizedAge =  mean(df$NormalizedAge),
                   NormalizedInterval =percRodCone, 
                   EntirePR = mean(df$EntirePR), 
                   SpecificPR = mean(df$SpecificPR))

dNormalizedInterval <- rbind(dConePR, dRodPR, dRodConePR)
# Store the predicted probabilities for each value of RodCone and D50 and append as last 3 columns to dataframe (columns 9, 10, 11)
pp <- cbind(dNormalizedInterval, predict(mdl, newdata = dNormalizedInterval, type = "probs", se = TRUE))

# calculate the mean probabilities within each level of RodCone 
#by(pp[, (ncol(pp)-2):ncol(pp)], pp$RodCone, colMeans)

# Collapse the data frame by outcome
lpp <- melt(pp, id.vars = c("RodCone", "NormalizedInterval"), measure.vars = levels(df$Outcome), value.name = "Probability")

# Plot predicted probabilities across NormalizedInterval values for each level of RodCone facetted by Outcome
ggplot(lpp, aes(x = NormalizedInterval, y = Probability, colour = RodCone)) + 
  geom_line() + facet_grid(variable ~ ., scales = "free_x") + 
  ggtitle("Predicted Outcome by Interval (normalized by developmental time)") + 
  theme(plot.title = element_text(hjust = 0.5)) + ylim(0,1)

# Plot predicted probabilities across NormalizedInterval values for each level of Oucome facetted by RodCone
ggplot(lpp, aes(x = NormalizedInterval, y = Probability, colour = variable)) + 
  geom_line() + facet_grid(RodCone ~ ., scales = "free_x") + 
  ggtitle("Predicted Outcome by Interval (normalized by developmental time)") + 
  theme(plot.title = element_text(hjust = 0.5)) + ylim(0,1)
```

### Entire PR

Predict probability by total amount of photoreceptors affected, assuming the following distribution:
Cones: between 0 and 3%, Rods: between 0 and 97%

```{R Predicted Prob Entire PR model 3}
percCone = seq(0, 3, by=0.5)
percRod = seq(0, 97, by=0.5)
percRodCone = seq(0, 100, by=0.5)

dConePR <-data.frame(RodCone = rep("Cone", each = length(percCone)), 
                   NormalizedAge =  mean(df$NormalizedAge),
                   NormalizedInterval = mean(df$NormalizedInterval), 
                   EntirePR = percCone, 
                   SpecificPR = mean(df$SpecificPR))

dRodPR <-data.frame(RodCone = rep("Rod", each = length(percRod)), 
                   NormalizedAge =  mean(df$NormalizedAge),
                   NormalizedInterval = mean(df$NormalizedInterval), 
                   EntirePR = percRod, 
                   SpecificPR = mean(df$SpecificPR))
  
dRodConePR <- data.frame(RodCone = rep("RodCone", each = length(percRodCone)), 
                   NormalizedAge =  mean(df$NormalizedAge),
                   NormalizedInterval = mean(df$NormalizedInterval), 
                   EntirePR = percRodCone, 
                   SpecificPR = mean(df$SpecificPR))

dEntirePR <- rbind(dConePR, dRodPR, dRodConePR)

# Store the predicted probabilities for each value of RodCone and D50 and append as last 3 columns to dataframe (columns 9, 10, 11)
pp <- cbind(dEntirePR, predict(mdl, newdata = dEntirePR, type = "probs", se = TRUE))

# calculate the mean probabilities within each level of RodCone 
#by(pp[, (ncol(pp)-2):ncol(pp)], pp$RodCone, colMeans)

# Collapse the data frame by outcome
lpp <- melt(pp, id.vars = c("RodCone", "EntirePR"), measure.vars = levels(df$Outcome), value.name = "Probability")

# Plot predicted probabilities across EntirePR values for each level of RodCone facetted by Outcome
ggplot(lpp, aes(x = EntirePR, y = Probability, colour = RodCone)) + 
  geom_line() + facet_grid(variable ~ ., scales = "free")  + 
  ggtitle("Predicted Outcome by Percentage of Enrire Photoreceptors Affected") + 
  theme(plot.title = element_text(hjust = 0.5))  + ylim(0,1)

# Plot predicted probabilities across EntirePR values for each level of Oucome facetted by RodCone
ggplot(lpp, aes(x = EntirePR, y = Probability, colour = variable)) + 
  geom_line() + facet_grid(RodCone ~ ., scales = "free")  + 
  ggtitle("Predicted Outcome by Percentage of Enrire Photoreceptors Affected") + 
  theme(plot.title = element_text(hjust = 0.5))  + ylim(0,1)
```

### Specific PR

Predict probability of each outcome by the amount of scpecific photoreceptors affected.

```{R Predicted Prob Specific PR model 3}
percRodCone = seq(0, 100, by=0.5)
typePR = c("Rod", "Cone")

dSpecificPR <- data.frame(RodCone = rep(typePR, each = length(percRodCone)), 
                   NormalizedAge =  mean(df$NormalizedAge),
                   NormalizedInterval = mean(df$NormalizedInterval), 
                   EntirePR =  mean(df$EntirePR),
                   SpecificPR = rep(percRodCone, length(typePR)))

# Store the predicted probabilities for each value of RodCone and D50 and append as last 3 columns to dataframe (columns 9, 10, 11)
pp <- cbind(dSpecificPR, predict(mdl, newdata = dSpecificPR, type = "probs", se = TRUE))

# calculate the mean probabilities within each level of RodCone 
#by(pp[, (ncol(pp)-2):ncol(pp)], pp$RodCone, colMeans)

# Collapse the data frame by outcome
lpp <- melt(pp, id.vars = c("RodCone", "SpecificPR"), measure.vars = levels(df$Outcome), value.name = "Probability")

# Plot predicted probabilities across SpecificPR values for each level of RodCone facetted by Outcome
ggplot(lpp, aes(x = SpecificPR, y = Probability, colour = RodCone)) + 
  geom_line() + facet_grid(variable ~ ., scales = "free")  + 
  ggtitle("Predicted Outcome by Percentage of Specific Photoreceptors Affected") + 
  theme(plot.title = element_text(hjust = 0.5))  + ylim(0,1)

# Plot predicted probabilities across SpecificPR values for each level of Oucome facetted by RodCone
ggplot(lpp, aes(x = SpecificPR, y = Probability, colour = variable)) + 
  geom_line() + facet_grid(RodCone ~ ., scales = "free")  + 
  ggtitle("Predicted Outcome by Percentage of Specific Photoreceptors Affected") + 
  theme(plot.title = element_text(hjust = 0.5))  + ylim(0,1)
```

# Model 4: Manipulation
## Multinomial Regression Model

Multinomial regression (nnet:multinom), it is a log-linear model for which the resulting coefficients reflect the ratio of probability
Removed the Manipulation parameter because not significant in the model and the Method of investigtion to simplify prediction

```{r model 4}

df$Outcome<-relevel(df$Outcome, ref="Stability")

mdl <- multinom(Outcome ~ Manipulation, data = df)
summary(mdl)
```

## ANOVA

Parameters significant for the model are: Rod/Cone affected and Entire PR affected.

```{R anova model 4}
Anova(mdl)
```
## Wald-test (2-tailed z test)
```{r Z-test model 4}
z <- summary(mdl)$coefficients/summary(mdl)$standard.errors
p <- (1 - pnorm(abs(z), 0, 1)) * 2
p
```

## Relative risk

```{r extract coefficients model 4}
writeLines("Ratio of the probability of choosing one outcome category over the reference category.\nOften referred as relative risk (and sometimes referred to as odds).\nCalculated by extracting the coefficients from the model and exponentiating\n")
exp(coef(mdl))

#writeLines("\nPredicted probabilities for the source data (head of table)")
#head(fitted(mdl))
```

## Predicted probabilities
### Categorical Parameters

```{r predicted probabilities model 4, results='asis'}
dsim <- data.frame(Manipulation = levels(df$Manipulation))
pred <- predict(mdl, newdata = dsim, "probs")

dfPred <- as.data.frame(pred)
row.names(dfPred) <- NULL
dfPred <- roundDf(dfPred, digits=2)
Manipulation <- c(levels(df$Manipulation)) 
dfPred <- cbind(Manipulation, dfPred)
z <- ztable(dfPred) %>%
  print(caption = "Predicted utcome probability by manipulation")
```

# Session Info
```{r session info}
sessionInfo()
```